---
title: "Introduction to analogs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to analogs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
fig.width = 7,
fig.height = 5
)
```

```{r setup}
library(analogs)
library(terra)
```

## Overview

The **analogs** package implements a general framework for distance-based neighborhood models. Every analysis follows a two-stage pattern:

1. First, **select** a neighborhood of analog locations from a reference pool, based on climate similarity, geographic proximity, or both.
2. Optionally, **summarize** the neighborhood using counts, weighted means, regression coefficients, or other statistics.

Common methods in climate change ecology---climate velocity, analog impact models, analog availability---are all specific configurations of this framework. So are geographically weighted regression (GWR) and inverse distance-weighted (IDW) interpolation, which use purely geographic neighborhoods. The package implements these via a flexible core function, `analog_search()`, along with simplified wrappers for common analysis types.

The table below shows how each wrapper maps to the framework:

| Function | Neighborhood | Summary | Use case |
|---|---|---|---|
| `analog_velocity()` | k nearest geographic, climate-constrained | Pairs | Where is this climate moving? |
| `analog_similarity()` | k nearest climatic, geographically-constrained | Pairs | What climates are reachable? |
| `analog_availability()` | All within thresholds | Count | Where do analogs exist? |
| `analog_intensity()` | All within thresholds | Weighted sum | How strong are analog matches? |
| `analog_impact()` | All within thresholds | Weighted mean | What ecological conditions to expect? |
| `analog_regression()` | Flexible | Local regression | How do covariates predict outcomes within neighborhoods? |


## Data

All functions in the package require two data inputs: `x` (the set of focal locations for which analogs are to be found) and `pool` (the set of reference locations to search for analogs). These can be provided as matrices, data.frames, or SpatRasters. `x` and `pool` must have the same set of variables, but can be in different formats and can represent different sets of sites. 

Most of the examples in this vignette are focused on climate change, and assume that `x` and `pool` come from different time periods with different climates. In these situations, you typically have a choice about which dataset to map to which of the two parameters, and swapping them can give different insights. For example, as shown in the climate velocity section below, swapping the parameter mapping switches the analysis from forward to reverse velocity; while it's only shown for velocity, this is also possible for the other climate change analyses. For non-climate-change applications like GWR and IDW shown toward the end of this vignette, `x` and `pool` can be from the same time period.

The package includes an example dataset of climate rasters covering a portion of western North America at approximately 5 km resolution, with two scaled climate variables (climatic water deficit and actual evapotranspiration) for a historic baseline and a future projection (2041--2070, SSP585). These are used throughout this vignette.

```{r data}
clim <- example_rasters()
hist <- clim$historic
fut <- clim$future

plot(hist, main = c("Historic CWD", "Historic AET"))
```


## Climate velocity

Analog-based climate velocity metrics are widely used to assess dispersal constraints under climate change. Forward or outbound velocity, the distance from a site to the nearest location with a projected future climate similar to the site's historical/current climate, represents the migration required for current residents to track baseline climate conditions. Reverse or inbound velocity, the distance from a site to the nearest location with a historic climate similar to the site's projected future climate, represents constraints on the arrival of new residents adapted to the site's future conditions. 

These metrics are implemented as a geographic nearest-neighbor search under a hard climate constraint, with reverse velocity implemented by swapping the climate data for the two eras. For each focal site, the result returns data on its single nearest geographic analog (subject to the climate similarity threshold `max_clim`). The geographic distance to that analog, divided by the time elapsed, is the velocity. Sites with no analog in the reference pool have `NA` results. 

```{r velocity}
fwd_vel <- analog_velocity(
      x = hist,
      pool = fut,
      max_clim = 0.25,
      k = 1
)

rev_vel <- analog_velocity(
      x = fut,
      pool = hist,
      max_clim = 0.25,
      k = 1
)

plot(c(fwd_vel$geog_dist, rev_vel$geog_dist), 
     main = c("Forward velocity (km)", "Reverse velocity (km)"))
```

In addition to geographic and climatic distances, the result also includes the spatial coordinates of each site's nearest analog (as well as the analog's index in the `pool`). This can be used for a variety of downstream analyses, including visualizing the direction from a site to its analog location:

```{r bearing}
fwd_vel$bearing <- geosphere::bearing(crds(fwd_vel, na.rm = FALSE), 
                                      values(fwd_vel)[, c("analog_x", "analog_y")])

plot(fwd_vel$bearing, main = c("Bearing to forward analog (deg)"),
     col = rainbow(12))
```


## Climate similarity

Climate similarity answers the inverse question from climate velocity: for each location's climate, what is the most similar climate within a given geographic radius? This finds the most climatically similar locations that are geographically reachable, which can provide insight into how much climate change organisms must tolerate given limited dispersal capacity.

```{r similarity}
sim <- analog_similarity(
      x = fut,
      pool = hist,
      max_geog = 100,
      k = 1
)

plot(sim$clim_dist, main = "Similarity of best nearby climate analog")
```


## Analog availability

How many analogs exist within specified climate and geographic thresholds? This maps the density of suitable analogs across the landscape. Locations with large numbers of nearby analogs have greater potential for adaptive dispersal under climate change, while locations with zero analogs represent novel or disappearing climates.

```{r availability}
avail <- analog_availability(
      x = fut,
      pool = hist,
      max_clim = 0.5,
      max_geog = 200
)

plot(avail, main = "Analog availability (count)")
```


## Analog intensity

Similar to availability, but weighted by climate similarity and/or geographic proximity rather than simply counted. This captures both the number and quality of analog matches.

```{r intensity}
intens <- analog_intensity(
      x = fut,
      pool = hist,
      max_clim = 0.5,
      max_geog = 200,
      weight = "gaussian_clim",
      theta = 0.15
)

plot(intens, main = "Analog intensity")
```


## Analog impact models

The analog impact model (AIM) predicts ecological outcomes under climate
change. For each focal location's future climate, it finds current locations
with similar climates within dispersal range, then computes a weighted
average of their ecological characteristics. Here we use CWD values from
the historic period as a stand-in for an ecological response variable.

```{r impact}
# Use historic CWD as a proxy ecological variable
eco_var <- hist$CWD

impact <- analog_impact(
      x = fut,
      pool = hist,
      y = eco_var,
      max_clim = 0.5,
      max_geog = 200,
      weight = "gaussian_clim",
      theta = 0.15
)

plot(impact$weighted_mean, main = "Predicted ecological state (AIM)")
```

The `weight` and `theta` parameters control how analog influence decays with climate distance. A Gaussian kernel with `theta = 0.15` means analogs at a climate distance of 0.15 receive about 60% weight, while those at 0.45 (= 3 × theta) receive almost none. The hard threshold `max_clim` provides an absolute cutoff.


## Spatial interpolation

The analog framework is also useful for analyses outside climate change scenarios. When `x` and `pool` share the same climate era, `analog_impact()` performs climate- and/or geographically-informed interpolation: for each grid cell, it finds sample locations with similar climates or nearby locations, then computes a weighted average of their measured values. This is useful for mapping ecological variables from sparse field observations onto a continuous grid. Unlike purely geographic interpolation methods (inverse distance weighting, kriging), this approach can weight observations by climate similarity, so a distant site with matching climate can contribute more than a nearby site with different climate.

The example below shows an interpolation informed by both climatic similarity and geographic proximity, with their relative importance determined by the `theta` parameters and the shape of the selected `weight` function.


```{r interpolation}
# Simulate sparse field observations at 500 random locations
set.seed(123)
n_sites <- 500
cells <- sample(which(!is.na(values(hist[[1]]))), n_sites)
sites <- as.data.frame(hist, xy = TRUE, na.rm = FALSE)[cells, ]
sites$observed <- 2 * sites$CWD - 0.5 * sites$AET + rnorm(n_sites, sd = 0.3)

# Interpolate onto the full grid
interp <- analog_impact(
  x = hist,
  pool = as.matrix(sites[, c("x", "y", "CWD", "AET")]),
  y = sites$observed,
  max_clim = 0.5,
  max_geog = 200,
  weight = "gaussian_joint",
  theta = c(0.15, 100)
)

plot(interp[["weighted_mean"]], main = "Climate-informed interpolation")
# points(sites[, c("x", "y")], col = "red")
```


## Local regression

When auxiliary covariates are available at pool locations, `analog_regression()` fits a weighted least squares regression within each analog neighborhood and returns the coefficients. This generalizes the weighted mean by resolving covariate-driven variation. The slope coefficients capture how much the response varies with the covariates within each neighborhood.

```{r regression}
# Simulate covariates for the pool (in practice these would be real measurements)
set.seed(42)
n_pool <- ncell(hist)
pool_covariates <- data.frame(
     aet = as.vector(values(hist$AET)),
     aet2 = as.vector(values(hist$AET))^2
)

reg <- analog_regression(
      x = fut,
      pool = hist,
      y = eco_var,
      covariates = pool_covariates,
      max_clim = 0.5,
      max_geog = 200,
      weight = "gaussian_clim",
      theta = 0.15,
      lambda = 1
)

plot(reg$intercept, main = "Regression intercept")
plot(reg$aet, main = "Coefficient: cov1")
```

With `lambda = 0` (the default), this is ordinary weighted least squares. Setting `lambda > 0` adds ridge regularization, which shrinks covariate coefficients toward zero and the intercept toward the weighted mean. This is useful when some neighborhoods have few analogs relative to the number of covariates.


## Geographically weighted regression

The same regression machinery supports geographically weighted regression (GWR) by using geographic neighborhoods and geographic distance weighting. This is a different configuration of the same underlying framework — no climate constraint, geographic kernel, local regression on covariates.

```{r gwr}
# GWR: geographic neighborhood, geographic weighting
gwr <- analog_regression(
      x = hist,
      pool = hist,
      y = eco_var,
      covariates = pool_covariates,
      select = "all",
      max_geog = 100,
      max_clim = NULL,
      weight = "inverse_geog",
      theta = 30,
      lambda = 0
)

plot(gwr$intercept, main = "GWR intercept")
plot(gwr$cov1, main = "GWR coefficient: cov1")
```


## Distance metrics

### Geographic distance

The package automatically detects whether coordinates are longitude/latitude or projected, and uses great-circle or planar distance accordingly. You can override this with `coord_type = "lonlat"` or `coord_type = "projected"`. Geographic thresholds (`max_geog`) are in kilometers for lon/lat data and in projection units for projected data.

### Climate distance

By default, climate distance is Euclidean. When using multiple climate variables, it is recommended to scale them first to avoid artifacts from differing units. The example data is already scaled.

For dataset-wide Mahalanobis distance (accounting for covariance among climate variables), use `mahalanobis_transform()` to pre-whiten the data:

```{r mahalanobis, eval = FALSE}
transformed <- mahalanobis_transform(x = hist, pool = fut)
vel_mahal <- analog_velocity(
      x = transformed$x,
      pool = transformed$pool,
      max_clim = 2,
      k = 1
)
```

For site-specific covariance (e.g., based on local temporal climate variability), supply pre-computed covariance matrices via the `x_cov` parameter.


## Weighting functions

The `weight` parameter controls how analog influence decays with distance. Available options:

- `"uniform"` — all analogs weighted equally
- `"gaussian_clim"` / `"gaussian_geog"` — Gaussian kernel on climate or geographic distance
- `"inverse_clim"` / `"inverse_geog"` — inverse distance weighting
- `"gaussian_joint"` / `"inverse_joint"` — kernels on combined climate and geographic distance

The `theta` parameter controls the bandwidth. For Gaussian kernels, `theta` is the standard deviation of the kernel. A rule of thumb: set `theta` so that the weight decays to near zero at the `max_clim` or `max_geog` boundary (roughly `theta ≈ max / 3`).

For joint weights, `theta` is a 2-element vector `c(theta_clim, theta_geog)`.


## Computational performance

The package is designed for large-scale analyses involving millions of pairwise comparisons.

### Pre-built indices

When running multiple queries against the same reference pool, build the lattice index once and reuse it:

```{r index}
idx <- build_analog_index(hist)

# Multiple queries reuse the same index
avail_tight <- analog_availability(fut, idx, max_clim = 0.3, max_geog = 100)
avail_loose <- analog_availability(fut, idx, max_clim = 0.8, max_geog = 300)
```

### Index tuning

By default, the lattice resolution is auto-tuned for each query. For repeated queries with the same configuration, you can tune once and reuse:

```{r tune, eval = FALSE}
res <- tune_index_res(
      x = fut, pool = hist,
      stat = "count",
      max_clim = 0.5, max_geog = 200,
      verbose = TRUE
)
idx <- build_analog_index(hist, index_res = res)
```

### Parallel processing

Use the `n_threads` parameter to parallelize across focal locations:

```{r threads, eval = FALSE}
result <- analog_velocity(fut, hist, max_clim = 0.5, k = 1, n_threads = 4)
```

### Large raster datasets

For rasters too large to fit in memory, `tiled_analog_search()` processes the focal data in spatial tiles:

```{r tiled, eval = FALSE}
result <- tiled_analog_search(
      x = very_large_raster,
      pool = idx,
      stat = "count",
      max_clim = 0.5,
      max_geog = 200,
      n_tiles = 16
)
```

### Downsampling

For very large reference pools, downsampling reduces computation. The package uses an adaptive sampling routine that reduces the effects on precision by downsampling more heavily in dense regions of climatic-geographic space in order to preserve coverage in sparse regions:

```{r downsample, eval = FALSE}
result <- analog_availability(
      x = fut, pool = hist,
      max_clim = 0.5, max_geog = 200,
      downsample = 0.1
)
```
